---
title: "Master Data Analysis"
output:
  html_document: default
  pdf_document: default
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This document serves as the master file for data analysis of the replication study, *"How Quick Decisions Illuminate Moral Character."* The primary objective of this study is to closely mirror the methodology of the original research. Accordingly, the statistical analysis will replicate the 2x2 ANOVA used in the original study, enabling a direct assessment of the main effects of decision speed and moral condition on moral character evaluations, as well as their interaction.

------------------------------------------------------------------------

# Data Collection and Analysis

This replication study is part of a series of replication projects conducted in **PSYCH 201a / CSS 204**. Three independent teams replicated the study. To strengthen the confirmatory analysis, we merged the datasets from all three teams. The analysis will be based on both the combined dataset and the individual datasets specific to this team.

-   **Team 1:** Led by \_\_\_\_\_\_ [Link to repository: ]\
-   **Team 2:** Led by \_\_\_\_\_\_ [Link to repository: ]\
-   **Team 3:** Led by the authors of this document: Luna Bellitto, Erika Garza, Emily Han, and Asad Tariq [Link to repository: ]

------------------------------------------------------------------------

# Hypotheses

1.  **Individuals who make quick decisions about moral dilemmas will be perceived as having stronger moral convictions compared to those who deliberate longer.**

    -   A 2 (moral condition) x 2 (decision speed) ANOVA will be conducted to compare group means for moral character evaluations, examining the main effects and interaction effects of both factors.

2.  **Additional analyses will explore perceived certainty and emotional impulsivity as covariates, assessing how perceptions of these traits vary with decision speed.**\
    *(Note: This section is still in progress and will be completed by the end of the week.)*

------------------------------------------------------------------------

### Loading required library and data set

```{r}
library(tidyverse)
library(dplyr)
library(car)
library(emmeans)
library(effectsize)
library(lme4)
library(ggplot2)
library(lmerTest)
```

```{r}
# Load the dataset
dat <- read.csv("quick_decisions_combined.csv")  # Read the dataset into a data frame
```

```{r}
# Count how many trials of each group is kept for combined analysis
trial_count1 <- dat %>%
  group_by(group) %>% 
  summarize(row_count = n())


# View the counts
print(trial_count1)



# cleaned_data <- dat %>%
#   # Step 1: Filter out missing values
#   mutate(is_complete = complete.cases(.)) %>%
#   filter(complete.cases(.)) 
# 
# trial_count2 <- cleaned_data %>%
#   group_by(group) %>% 
#   summarize(row_count = n())
# 
# print(trial_count2)
```

## Implement Exclusion Criteria: For participant with incomplete answers and lack of variance


```{r}

# Process the dataset
cleaned_data <- dat %>%
  # Step 1: Filter out missing values
  mutate(is_complete = complete.cases(.)) %>%
  filter(complete.cases(.)) %>%
  
  # Step 2: Calculate variance for Justin and Nate
  rowwise() %>%
  mutate(
    justin_variance = var(c(2:11), na.rm = TRUE), # Variance for Justin
    nate_variance = var(c(12:22), na.rm = TRUE)          # Variance for Nate
  ) %>%
  ungroup() %>%
  
  # Step 3: Filter out participants based on variance criteria
  filter(!(justin_variance == 0 | nate_variance == 0)) %>%
  
  # Step 4: Drop intermediate columns used for exclusion
  select(-is_complete, -justin_variance, -nate_variance)

# Step 5: Calculate exclusion counts
  na_exclusions_count <- nrow(dat) - nrow(dat %>% filter(complete.cases(.))) # Count rows with missing values
  variance_exclusions_count <- nrow(dat %>% filter(complete.cases(.))) - nrow(cleaned_data) # Count participants excluded for lack of variance
  total_exclusions <- na_exclusions_count + variance_exclusions_count  # Calculate total exclusions

# Step 6: Create a summary table of exclusions
exclusions_summary <- tibble(
  Exclusion_Criteria = c(
    "Missing values (NA)",
    "Lack of variance",
    "Total exclusions"
  ),
  Count = c(
    na_exclusions_count,
    variance_exclusions_count,
    total_exclusions
  )
)

# Display the table
print(exclusions_summary)

```


```{r}

# Count how many trials of each group is kept for combined analysis
trial_count <- cleaned_data %>%
  group_by(group) %>% 
  summarize(row_count = n())

# Count how many trials is kept for each condition in combined analysis
condition_combined_counts <- cleaned_data %>%
  group_by(condition) %>%
  summarize(count = n())

# View the counts
print(trial_count)
print(condition_combined_counts)

```

## Manipulation Check for Combined Dataset

Groups the data by condition and provides a clear comparison of how the participants perceived Justin's decision speed and Nate's decision speed across the two conditions. (Note: the mean should be different between Justin and Nate but it should be similar across condition).

Then we will determine whether there is a statistically significant difference in the mean scores of Justin's speed and Nate's speed.

```{r}
# Add Participant_ID column and make it the first column
cleaned_data <- cleaned_data %>%
  mutate(Participant_ID = row_number()) %>%
  select(Participant_ID, everything())

```

```{r}
#Summarize the decision speed columns for comparison

manipulation_stats <- cleaned_data %>%
  group_by(condition) %>%  # Group data by the 'condition' column
  summarize(
    mean_speed_justin = mean(Justin_Q1),  # Mean of Justin's speed
    mean_speed_nate = mean(Nate_Q1)     # Mean of Nate's speed
  )


# Paired t-test for Justin's and Nate's speeds for checking whether the manipulation was successful or not 
manipulation_combined_t_test <- t.test(
  cleaned_data$Justin_Q1, 
  cleaned_data$Nate_Q1, 
  paired = TRUE
)

print(manipulation_combined_t_test)
```

### Perceived Decision Speed Across Condition [**Not Necessary**]

Upon reviewing the descriptive statistics using **'manipulation_stats'**, we observed a potential mean difference in perceived decision speed for Nate across conditions (moral vs. immoral). Out of curiosity, we decided to conduct a statistical test to explore whether this observed difference is significant.

Please note that this analysis is conducted purely for exploratory purposes and is not part of the original study design. The results should be interpreted cautiously and are not intended to inform the main findings of the study.

```{r}
# Independent t-tests for moral vs immoral condition on Justin's speed & Nate's speed 
t_test_justin <- t.test(
  Justin_Q1 ~ condition, 
  data = cleaned_data,
  var.equal = TRUE
)

t_test_nate <- t.test(
  Nate_Q1 ~ condition, 
  data = cleaned_data,
  var.equal = TRUE
)

print(t_test_justin)
print(t_test_nate)

```

## Confirmatory Analysis for Combined Dataset: 2x2 Anova

**Objective** : The aim of this analysis is to perform a **2 (moral condition: moral vs. immoral) x 2 (decision speed: quick vs. slow)** repeated-measures ANOVA to compare group means for moral character evaluations of Justin and Nate.

### Step 1: Calculate Average Scores for Justin and Nate

```{r}
character_eval_combined <- cleaned_data %>% 
  mutate(
    # Calculate the average score for Justin across selected questions
    quick_justin_score = rowMeans(select(., Justin_Q2, Justin_Q3, Justin_Q4)),
    # Calculate the average score for Nate across selected questions
    slow_nate_score = rowMeans(select(., Nate_Q2, Nate_Q3, Nate_Q4))
  )
```

### Step 2: Summarize Average Scores by Condition

Group data by condition for comparison of Justin's moral character evaluation and Nate's character evaluation across condition

**[Note for ERI]: I think you can use this to create the plot**

```{r}
 character_eval_combined_stats <- character_eval_combined %>%
  group_by(condition) %>%  # Group data by the 'condition' column
  summarize(
    mean_justin = mean(quick_justin_score), # Mean of Justin's character evaluation
    mean_nate = mean(slow_nate_score)     # Mean of Nate's character evaluation
  )

print(character_eval_combined_stats)  # View summarized data
```

### Step 3: Reshape Data for ANOVA

```{r}
character_eval_combined_long <- character_eval_combined %>%
  pivot_longer(
    cols = c(quick_justin_score, slow_nate_score), # Select the new columns
    names_to = "speed",                           # Create a column for speed
    values_to = "score"                            # Create a column for scores
  ) %>%
  mutate(
    speed = ifelse(speed == "quick_justin_score", "Justin", "Nate") # Relabel speed column
  )  %>%
  select(Participant_ID, condition, speed, score) # Keep only essential columns


# View the reshaped data
head(character_eval_combined_long)
```

### Step 4: Run Repeated-Measures ANOVA

```{r}
anova_combined <- aov(score ~ condition * speed + Error(Participant_ID/speed), data = character_eval_combined_long)

# View the summary
summary(anova_combined)
```

### Step 5: Conduct Post Hoc Pairwise Comparisons

```{r}
# Conduct post hoc pairwise comparisons using estimated marginal means
posthoc_results <- emmeans(anova_combined, pairwise ~ condition * speed)

print(posthoc_results)

```


### Step 6: Compute the effect size
```{r}
# Refit the repeated measures model
anova_combined_lmer <- lmer(
  score ~ condition * speed + (1 | Participant_ID),
  data = character_eval_combined_long
)

# Calculate partial eta-squared for the mixed-effects model
eta2_combined <- eta_squared(anova_combined_lmer, partial = TRUE)

# Print results
print(eta2_combined)
```

### Visualization

**Note to Eri: Please look at step 2**

```{r}

# Updated plot with detailed y-axis tweaks
ggplot(character_eval_combined_long, aes(x = speed, y = score, fill = speed)) +
  geom_bar(
    stat = "summary",
    fun = "mean",
    position = position_dodge(width = 0.5),  # Columns closer together
    width = 0.6                             # Narrower columns
  ) +
  facet_wrap(
    ~ condition, 
    labeller = as_labeller(c(moral = "Moral", immoral = "Immoral")),  # Simplify labels
    strip.position = "bottom"  # Moral/Immoral labels at the bottom
  ) +
  labs(
    title = "Moral Character Evaluations (Combined Data)",
    y = "Mean Score (Likert Scale)",  # Y-axis label with clarification
    fill = "Decision Speed"  # Legend title
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 12),
    strip.text = element_text(face = "bold"),
    strip.placement = "outside",  # Push bottom labels outward
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),  # Centered title
    panel.spacing = unit(1, "lines"),  # Adjust spacing between panels
    axis.title.x = element_blank(),  # Remove x-axis title
    axis.text.x = element_blank(),   # Remove x-axis labels (Justin/Nate)
    axis.title.y = element_text(margin = margin(r = 15))  # Add spacing to y-axis label
  ) +
  scale_fill_manual(
    values = c("Justin" = "#4F81BD", "Nate" = "#9DC3E6"),  # Two shades of blue
    labels = c("Justin" = "Quick (Justin)", "Nate" = "Slow (Nate)")  # Custom legend labels
  ) +
  scale_y_continuous(
    limits = c(0, 7),  # Y-axis limits for Likert scale
    breaks = 1:7,      # Label y-axis by 1's
    expand = c(0, 0)   # No extra space above/below
  )


```

```{r}
ggplot(character_eval_combined_long, aes(x = speed, y = score, fill = condition)) +
  geom_bar(
    stat = "summary",
    fun = "mean",
    position = position_dodge(width = 0.6),  # Columns closer together
    width = 0.5                             # Narrower columns
  ) +
  labs(
    title = "Moral Character Evaluations (Combined Data)",
    x = "Decision Speed (Justin / Nate)",
    y = "Mean Score (Likert Scale)",  # Updated y-axis label
    fill = "Moral Condition"          # Legend title for conditions
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 12),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),  # Centered title
    axis.text.x = element_text(size = 10),   # Adjust x-axis labels
    axis.title.y = element_text(size = 12, margin = margin(r = 15)),  # Add spacing to y-axis title
    axis.title.x = element_text(size = 12, margin = margin(t = 10))   # Add spacing to x-axis title
  ) +
  scale_fill_manual(
    values = c("moral" = "#4F81BD", "immoral" = "#9DC3E6"),  # Two shades for moral/immoral
    labels = c("moral" = "Moral", "immoral" = "Immoral")     # Custom legend labels
  ) +
  scale_y_continuous(
    limits = c(0, 7),  # Y-axis limits for Likert scale
    breaks = 1:7,      # Label y-axis by 1's
    expand = c(0, 0)   # No extra space above/below
  )
```


```{r}

# Step 1: Summarize the data to get means for each speed and condition
line_data <- character_eval_combined_long %>%
  group_by(speed, condition) %>%
  summarize(mean_score = mean(score), .groups = "drop")

# Step 2: Plot with connecting lines and tweaks
ggplot(character_eval_combined_long, aes(x = condition, y = score, fill = speed)) +
  # Bar plot for the data
  geom_bar(
    stat = "summary",
    fun = "mean",
    position = position_dodge(width = 0.8),  # Columns closer together
    width = 0.6                             # Narrower columns
  ) +
  # Add lines connecting the means
  geom_line(
    data = line_data,
    aes(x = as.numeric(as.factor(condition)) + ifelse(speed == "Justin", -0.2, 0.2), 
        y = mean_score, group = speed),
    size = 1.3,  # Line thickness
    color = "darkblue"
  ) +
  # Add dots at the ends of the lines, centered with the bars
  geom_point(
    data = line_data,
    aes(x = as.numeric(as.factor(condition)) + ifelse(speed == "Justin", -0.2, 0.2), 
        y = mean_score),
    size = 3,
    color = "darkblue"
  ) +
  labs(
    title = "Moral Character Evaluations Combined Data",
    x = "Condition",
    y = "Mean Score (Likert Scale)",  # Y-axis label with clarification
    fill = "Decision Speed"
  ) +
  scale_fill_manual(
    values = c("Justin" = "#4F81BD", "Nate" = "#9DC3E6"),  # Keep column colors the same
    labels = c("Justin" = "Quick (Justin)", "Nate" = "Slow (Nate)")  # Custom legend labels
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 12),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),  # Centered title
    axis.text.x = element_text(size = 10),
    axis.title.y = element_text(margin = margin(r = 15)),  # Add spacing to y-axis label
    legend.position = "none"  # Remove the legend for speed (lines)
  ) +
  scale_y_continuous(
    limits = c(0, 7),  # Y-axis limits for Likert scale
    breaks = 1:7,      # Label y-axis by 1's
    expand = c(0, 0)   # No extra space above/below
  )

```

## Confirmatory Analysis of Group 3's data: 2x2 Anova

This analysis focuses on a subset of the data, specifically Group 3. We will perform the same 2 (moral condition: moral vs. immoral) x 2 (decision speed: quick vs. slow) ANOVA as in the full dataset but restricted to Group 3. This allows us to assess whether the observed effects are consistent within this specific subset.

### Sub setting the data

```{r}
g3_cleaned_data <- cleaned_data %>%
  filter(group == 3)
  
# Count how many trials is kept for each condition in combined analysis
g3_condition_counts <- g3_cleaned_data %>%
  group_by(condition) %>%
  summarize(count = n())

print(g3_condition_counts)

```

### Step 1: Calculate Average Scores for Justin and Nate

```{r}
character_eval_group3 <- g3_cleaned_data %>% 
  mutate(
    # Calculate the average score for Justin across selected questions
    quick_justin_score = rowMeans(select(., Justin_Q2, Justin_Q3, Justin_Q4)),
    # Calculate the average score for Nate across selected questions
    slow_nate_score = rowMeans(select(., Nate_Q2, Nate_Q3, Nate_Q4))
  )
```

### Step 2: Summarize Average Scores by Condition

Group data by condition for comparison of Justin's moral character evaluation and Nate's character evaluation across condition

**[Note for ERI]: I think you can use this to creat the plot**

```{r}
 character_eval_group3_stats <- character_eval_group3 %>%
  group_by(condition) %>%  # Group data by the 'condition' column
  summarize(
    mean_justin = mean(quick_justin_score), # Mean of Justin's character evaluation
    mean_nate = mean(slow_nate_score)     # Mean of Nate's character evaluation
  )

print(character_eval_combined_stats)  # View summarized data
```

### Step 3: Reshape Data for ANOVA

```{r}
character_eval_group3_long <- character_eval_group3 %>%
  pivot_longer(
    cols = c(quick_justin_score, slow_nate_score), # Select the new columns
    names_to = "speed",                           # Create a column for speed
    values_to = "score"                            # Create a column for scores
  ) %>%
  mutate(
    speed = ifelse(speed == "quick_justin_score", "Justin", "Nate") # Relabel speed column
  )  %>%
  select(Participant_ID, condition, speed, score) # Keep only essential columns


# View the reshaped data
head(character_eval_group3_long)
```

### Step 4: Run Repeated-Measures ANOVA

```{r}
anova_group3 <- aov(score ~ condition * speed + Error(Participant_ID/speed), data = character_eval_group3_long)

# View the summary
summary(anova_group3)
```

### Step 5: Conduct Post Hoc Pairwise Comparisons

```{r}
# Conduct post hoc pairwise comparisons using estimated marginal means
posthoc_group3_results <- emmeans(anova_group3, pairwise ~ condition * speed)

print(posthoc_group3_results)

```

### Step 6: Compute the effect size
```{r}
# Refit the repeated measures model
anova_group3_lmer <- lmer(
  score ~ condition * speed + (1 | Participant_ID),
  data = character_eval_group3_long
)

# Calculate partial eta-squared for the mixed-effects model
eta2_group3 <- eta_squared(anova_group3_lmer, partial = TRUE)

# Print results
print(eta2_group3)
```

### Visualization

**Note to Eri: Please look at step 2**

```{r}

# Updated plot with detailed y-axis tweaks
ggplot(character_eval_group3_long, aes(x = speed, y = score, fill = speed)) +
  geom_bar(
    stat = "summary",
    fun = "mean",
    position = position_dodge(width = 0.5),  # Columns closer together
    width = 0.6                             # Narrower columns
  ) +
  facet_wrap(
    ~ condition, 
    labeller = as_labeller(c(moral = "Moral", immoral = "Immoral")),  # Simplify labels
    strip.position = "bottom"  # Moral/Immoral labels at the bottom
  ) +
  labs(
    title = "Moral Character Evaluations (Group 3 Data)",
    y = "Mean Score (Likert Scale)",  # Y-axis label with clarification
    fill = "Decision Speed"  # Legend title
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 12),
    strip.text = element_text(face = "bold"),
    strip.placement = "outside",  # Push bottom labels outward
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),  # Centered title
    panel.spacing = unit(1, "lines"),  # Adjust spacing between panels
    axis.title.x = element_blank(),  # Remove x-axis title
    axis.text.x = element_blank(),   # Remove x-axis labels (Justin/Nate)
    axis.title.y = element_text(margin = margin(r = 15))  # Add spacing to y-axis label
  ) +
  scale_fill_manual(
    values = c("Justin" = "#4F81BD", "Nate" = "#9DC3E6"),  # Two shades of blue
    labels = c("Justin" = "Quick (Justin)", "Nate" = "Slow (Nate)")  # Custom legend labels
  ) +
  scale_y_continuous(
    limits = c(0, 7),  # Y-axis limits for Likert scale
    breaks = 1:7,      # Label y-axis by 1's
    expand = c(0, 0)   # No extra space above/below
  )


```

```{r}

# Step 1: Summarize the data to get means for each speed and condition
line_data <- character_eval_group3_long %>%
  group_by(speed, condition) %>%
  summarize(mean_score = mean(score), .groups = "drop")

# Step 2: Plot with connecting lines and tweaks
ggplot(character_eval_group3_long, aes(x = condition, y = score, fill = speed)) +
  # Bar plot for the data
  geom_bar(
    stat = "summary",
    fun = "mean",
    position = position_dodge(width = 0.8),  # Columns closer together
    width = 0.6                             # Narrower columns
  ) +
  # Add lines connecting the means
  geom_line(
    data = line_data,
    aes(x = as.numeric(as.factor(condition)) + ifelse(speed == "Justin", -0.2, 0.2), 
        y = mean_score, group = speed),
    size = 1.3,  # Line thickness
    color = "darkblue"
  ) +
  # Add dots at the ends of the lines, centered with the bars
  geom_point(
    data = line_data,
    aes(x = as.numeric(as.factor(condition)) + ifelse(speed == "Justin", -0.2, 0.2), 
        y = mean_score),
    size = 3,
    color = "darkblue"
  ) +
  labs(
    title = "Moral Character Evaluations (Group 3 Data)",
    x = "Condition",
    y = "Mean Score (Likert Scale)",  # Y-axis label with clarification
    fill = "Decision Speed"
  ) +
  scale_fill_manual(
    values = c("Justin" = "#4F81BD", "Nate" = "#9DC3E6"),  # Keep column colors the same
    labels = c("Justin" = "Quick (Justin)", "Nate" = "Slow (Nate)")  # Custom legend labels
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 12),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),  # Centered title
    axis.text.x = element_text(size = 10),
    axis.title.y = element_text(margin = margin(r = 15)),  # Add spacing to y-axis label
    legend.position = "none"  # Remove the legend for speed (lines)
  ) +
  scale_y_continuous(
    limits = c(0, 7),  # Y-axis limits for Likert scale
    breaks = 1:7,      # Label y-axis by 1's
    expand = c(0, 0)   # No extra space above/below
  )

```
```{r}
ggplot(character_eval_group3_long, aes(x = speed, y = score, fill = condition)) +
  geom_bar(
    stat = "summary",
    fun = "mean",
    position = position_dodge(width = 0.6),  # Columns closer together
    width = 0.5                             # Narrower columns
  ) +
  labs(
    title = "Moral Character Evaluations (Group 3 Data)",
    x = "Decision Speed (Justin / Nate)",    # Updated x-axis label
    y = "Mean Score (Likert Scale)",         # Updated y-axis label
    fill = "Moral Condition"                # Legend title for conditions
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 12),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),  # Centered title
    axis.text.x = element_text(size = 10),   # Adjust x-axis labels
    axis.title.y = element_text(size = 12, margin = margin(r = 15)),  # Add spacing to y-axis title
    axis.title.x = element_text(size = 12, margin = margin(t = 10))   # Add spacing to x-axis title
  ) +
  scale_fill_manual(
    values = c("moral" = "#4F81BD", "immoral" = "#9DC3E6"),  # Two shades for moral/immoral
    labels = c("moral" = "Moral", "immoral" = "Immoral")     # Custom legend labels
  ) +
  scale_y_continuous(
    limits = c(0, 7),  # Y-axis limits for Likert scale
    breaks = 1:7,      # Label y-axis by 1's
    expand = c(0, 0)   # No extra space above/below
  )
```

