```{r setup, include=FALSE}
# Define the date
date_format <- format(Sys.time(), '%B %d, %Y')
```

---
title: "Replication of Study 'How Quick Decisions Illuminate Moral Character' by Clayton R. Critcher, Yoel Inbar and David A. Pizarro (2013, Social Psychological and Personality Science)"
author: "Critcher_2013_3"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc_depth: 3
---

## Introduction

This report presents our replication of the study "How Quick Decisions Illuminate Moral Character" conducted by Critcher, Inbar, and Pizarro (2013), which examined the effect of decision speed on judgments of moral character. The original study found that individuals who made quick moral decisions were evaluated more positively, whereas those who made quick immoral decisions were judged more harshly. In contrast, slower decisions, whether moral or immoral, led to more moderate evaluations of character. The authors proposed that quick decisions are seen as more certain and, therefore, more revealing of the decision maker’s underlying motives.

The original study employed two experiments. The first involved participants evaluating two individuals’ moral character, one who made a quick decision and another who took longer to decide, in both ethical and unethical contexts. The second experiment explored a similar structure but focused on more complex moral dilemmas. Both experiments consistently showed that quick decisions amplified moral evaluations, with quick moral decisions being judged more favorably and quick immoral decisions being judged more harshly.

Our replication aimed to reproduce the findings of the first experiment by closely following the original methodology. Specifically, we focused on testing whether decision speed continued to amplify moral character judgments, with quick decisions leading to more polarized evaluations compared to slower ones. This replication project is part of a broader effort to assess the robustness of psychological research through reproducibility studies.

## Methods

### Power Analysis

In the absence of prior data or specific effect size estimates from our study to conduct a traditional power analysis, we followed standard practice and increased our original sample size of 119 by a factor of 2.5, setting a target of 298 participants. This approach was intended to ensure adequate power to detect effects across conditions, compensating for the lack of data.

### Planned Sample

For our replication experiment, we would require a sample size of 298 participants. In the original study, the participants were students from the University of California, Berkeley or members from the nearby community who were randomly assigned to different conditions to assess the influence of decision speed on moral evaluations. However, participants for this replication study were recruited through Prolific and screened to ensure they reside in the US and are fluent in English.

### Materials

All the materials for this replication study come from the original study, *How Quick Decisions Illuminate Moral Character*, conducted by Critcher, Inbar, and Pizarro (2013). Our paradigm can be found [here](https://ucsd-psych201a.github.io/critcher2013_3/).

#### Scenario:

Participants were randomly assigned to one of two conditions: moral or immoral. Within each condition, two agents, Justin and Nate, were presented, differing in their decision-making speed.

-   **Moral Condition**: In separate instances, Justin and Nate encounter a cash-filled wallet in a grocery store parking lot. Justin quickly decides to return the wallet to customer service rather than keeping the money. Nate ultimately makes the same decision, but only after a prolonged period of deliberation.

-   **Immoral Condition**: In separate instances, Justin and Nate find a cash-filled wallet in a grocery store parking lot. Justin quickly decides to take the money and leave. Nate reaches the same conclusion but only after taking a long time to decide.

#### Questionnaire:

Participants were asked to answer the following questions based on their evaluation of each agent, Justin and Nate. Each question corresponds to one of the categories below, with its own specific scale.

##### Quickness:

1.  Did \[name\] make his decision quickly or slowly?\
    **Scale:** 1 = particularly slowly, 7 = particularly quickly

##### Moral character evaluation:

2.  Regardless of \[name\]’s decision, does it sound like \[name\] has underlying moral principles that are good, bad, or somewhere in between?\
    **Scale:** 1 = completely bad, 4 = mixed, 7 = completely good

3.  Regardless of \[name\]’s decision, do you think \[name\] has moral standards that are good, bad, or somewhere in between?\
    **Scale:** 1 = completely bad, 4 = mixed, 7 = completely good

4.  Regardless of \[name\]’s decision, do you think \[name\] possesses the moral knowledge and principles necessary to do ‘the right thing’?\
    **Scale:** 1 = not at all, 4 = somewhat, 7 = completely

##### Certainty:

5.  Would you say \[name\] was quite certain in his decision, or did \[name\] have hesitations about his decision?\
    **Scale:** 1 = completely certain, 7 = considerable hesitations

6.  How close do you think \[name\] was to choosing the alternate course of action?\
    **Scale:** 1 = very close to, 7 = not close at all

7.  How conflicted do you think \[name\] felt in making the decision?\
    **Scale:** 1 = very conflicted, 7 = not at all conflicted

8.  Based on the information provided, do you think \[name\] had many reservations about the decision?\
    **Scale:** 1 = none at all, 7 = a whole lot

##### Emotional impulsivity:

9.  Do you think \[name\] was calm and emotionally contained while making the decision?\
    **Scale:** 1 = not at all, 7 = entirely so

10. To what extent do you think \[name\] became upset and acted without thinking?\
    **Scale:** 1 = not at all, 7 = entirely so

### Procedure

As in the original experiment (Critcher et al., 2013), our paradigm followed this procedure: “Participants read about Justin and Nate, two men who each independently came upon cash-filled wallets in the parking lot of a local grocery store". Justin ‘was able to decide quickly’ what to do, while Nate ‘was only able to decide after long and careful deliberation.’ Participants assigned to the moral condition learned that both men ‘did not steal the money but instead left the wallet with customer service.’ Those in the immoral condition learned that both men ‘pocketed the money and drove off.’”

Adhering to the procedures outlined by Critcher et al. (2013), after reading the scenario of Justin and Nate’s actions, participants completed four Likert-scale assessments, each scaled from 1 to 7. The first assessment evaluated the quickness of the actors’ decisions (manipulation check). The second assessed Justin and Nate’s morality, with items such as: “has entirely good (vs. entirely bad) moral principles,” “has good (vs. bad) moral standards,” and “deep down has the moral principles and knowledge to do the right thing.” The third assessment included four items evaluating the certainty of the actors’ decisions, such as: “how conflicted \[each\] felt when making his decision,” “how many reservations \[each\] had,” “was quite certain in his decision,” and “how far \[each\] was from choosing the alternate course of action.” Finally, the fourth assessment evaluated perceived emotional impulsivity, asking whether the actors were “calm and emotionally contained” or “upset and acted without thinking.”

### Design Overview

#### Factors

In this replication study, two factors were manipulated: participants were randomly assigned to a condition (moral vs. immoral), and within each condition, they were exposed to both scenarios where one actor made the decision quickly (moral or immoral), and the other made the same decision slowly.

#### Primary Measures

Four primary measures were collected from participants:

1.  **Quickness of Decision**: A manipulation check assessing participants’ perceptions of how quickly Justin and Nate made their decisions (e.g., “Did \[name\] make his decision quickly or slowly?”).

2.  **Moral Character Evaluation**: Participants rated each agent’s moral principles and standards using items such as, “Does \[name\] have underlying moral principles that are good, bad, or somewhere in between?” and “Does \[name\] possess the moral knowledge and principles necessary to do the right thing?”

3.  **Certainty**: Four items evaluated participants’ perceptions of each agent’s certainty in their decision, including “How conflicted \[name\] felt when making the decision” and “How close \[name\] was to choosing the alternate course of action.”

4.  **Emotional Impulsivity**: Two items assessed perceptions of each agent’s emotional impulsivity, such as “Was \[name\] calm and emotionally contained?” and “Did \[name\] become upset and act without thinking?”

#### Mixed Design

The replication study reflects the original study’s mixed design:

-   **Between-Subjects Factor**:
    -   **Moral Condition (moral vs. immoral)**: Participants were randomly assigned to one of two conditions—moral (where both characters return the wallet) or immoral (where both characters keep the wallet). Each participant experienced only one of these conditions, making this factor between-subjects.
-   **Within-Subjects Factor**:
    -   **Decision Speed (quick vs. deliberative)**: Each participant evaluated both Justin (quick decision) and Nate (deliberative decision). Because every participant observed and evaluated both decision speeds within their assigned condition, this factor is within-subjects.

#### Impact of Switching Between- and Within-Subjects Designs

Although this replication followed the original study’s mixed design, switching both factors (decision and speed) to either a fully within-subjects or fully between-subjects design is a possible choice.

A fully within-subjects design could increase statistical power and require fewer participants, as each person would experience all combinations of conditions (quick/slow and moral/immoral). This design reduces variability due to individual differences, making it easier to detect effects. However, it could also risk carryover effects, where experiencing one condition might influence responses in the next, potentially producing biased results. Additionally, participants may guess the study’s purpose and alter their responses to appear consistent or meet the experiment’s expectations.

In contrast, in a fully between-subjects design, each participant would experience only one combination of decision speed and moral condition, requiring a larger sample size to achieve reliable effects. With only one condition per participant, variability due to individual differences is higher, which reduces statistical power and sensitivity to detect effects. However, this design minimizes potential biases, such as carryover effects and demand characteristics, as participants encounter only one scenario.

#### Potential Confounding Variables

Potential confounding variables in this study include but are not limited to participants’ own biases and preconceptions about the decision-making process, which can manifest as personal perceptions of quick versus slow decisions and overall moral judgments. Another important confound relates to preconceptions about men’s moral character, as both characters in the scenario are male, and participants may carry implicit gender biases that influence their evaluations. Additionally, cultural perspectives, participants’ emotional states and contexts, as well as the moral complexity of the scenarios, may further impact these judgments.

### Differences from Original Study

This replication differs from the original study primarily in the participant sample and mode of data collection. Participants for this replication were recruited through the online platform Prolific, rather than from a local population at a university. Consequently, this replication was conducted entirely online and targeted a sample size of 300 participants, larger than the original sample of 119. Additionally, this replication focused exclusively on Experiment 1 from the original study and did not include subsequent experiments.

### Methods Addendum (Post Data Collection)

#### Actual Sample

For our replication experiment, we would require at least a sample size of 298 participants. In the original study, the 119 participants were students from the University of California, Berkeley or members from the nearby community who were randomly assigned to different conditions to assess the influence of decision speed on moral evaluations. However, participants for this replication study will be recruited through the data collection platform, Prolific, and only screened for residing in the US and being fluent in English.

#### Differences from pre-data collection methods plan

Within the CSS 204 course, three independent teams worked on replicating Critcher, Inbar, and Pizarro’s study, each conducting separate data collection efforts. Combined, these efforts aligned with the target sample size of 300 specified in our planned power analysis (rounded from 298). To facilitate the analysis, our Teaching Assistant, Janna Wennberg, merged the cleaned datasets from all three teams. She carefully organized the data to ensure consistency and accuracy for our individual analyses. This combined dataset provided greater statistical power and a broader participant base, while still allowing for comparisons between the combined dataset and the individual datasets collected by our team.

For this analysis, we used both the merged dataset and the individual dataset specific to our group (group 3) to run our confirmatory analysis: 2x2 Anova. Additionally, we reverse-coded items related to decision certainty (from our combined dataset) to align with the original study’s methodology. These steps were implemented to ensure consistency across datasets and preserve the validity of our findings.

## Results

### Data preparation

Our replication of Experiment 1 will closely mirror the original study’s 2x2 factorial design, manipulating moral condition (stealing vs. not stealing) and decision speed (quick vs. slow). The dependent variable will be the moral character evaluation, assessed using a Likert scale. As in the original study, this replication will also examine perceived certainty of the actor in the scenario as covariates. The questionnaire items for the covariates will also be assessed using a Likert scale.

#### Raw Data Processing

```{r, echo = TRUE, message = FALSE, warning = FALSE}

#### Load necessary libraries
library(tidyverse)
library(dplyr)
library(car)
library(emmeans)
library(effectsize)
library(lme4)
library(ggplot2)
library(lmerTest)

# Define the directory containing your CSV files
directory_path <- "../data/final-data"

# Initialize an empty tibble to store all cleaned data
cleaned_data <- tibble()

# Get a list of all CSV files in the directory
file_list <- list.files(path = directory_path, pattern = "\\.csv$", full.names = TRUE)

# Loop through each file in the directory
for (file in file_list) {
  # Read the CSV file
  data <- read_csv(file)
  
  # Flag to track if the attention check passes
  attention_check_passed <- FALSE
  
  # Check attention check questions
  attention_check_row <- data %>%
    filter(trial_type == "survey-html-form")
  
  # Validate attention check
  if (nrow(attention_check_row) > 0) {
    # Parse the JSON string in the response column
    attention_responses <- jsonlite::fromJSON(attention_check_row$response)
    
    # Define correct answers for attention check
    correct_answers <- list(
      "q1" = "Justin and Nate",
      "q2" = "A wallet"
    )
    
    # Check if all attention check answers are correct
    attention_check_passed <- all(
      attention_responses$q1 == correct_answers$q1 &&
        attention_responses$q2 == correct_answers$q2
    )
  }
  
  # Process data only if attention check passes
  if (attention_check_passed) {
    # Select relevant columns (do not drop NAs here)
    temp_cleaned <- data %>%
      select(condition, scenario, starts_with("response_Q"))
    
    # Determine who came first based on the first non-NA scenario entry
    temp_cleaned <- temp_cleaned %>%
      group_by(condition) %>%
      mutate(
        first_scenario = first(na.omit(scenario))  # Identify the first scenario for each participant
      ) %>%
      ungroup()
    
    # Convert condition to a factor for easier interpretation
    temp_cleaned <- temp_cleaned %>%
      mutate(
        condition = factor(condition, levels = c(0, 1), labels = c("immoral", "moral")),
        scenario = factor(scenario)
      ) %>%
      # Rearrange columns to place first_scenario after condition
      select(condition, first_scenario, scenario, starts_with("response_Q"))
    
    # Rename response columns to include scenario information
    temp_cleaned <- temp_cleaned %>%
      # Ensure no NA in 'scenario' or 'question' before combining
      filter(!is.na(scenario)) %>%
      pivot_longer(cols = starts_with("response_Q"),
                   names_to = "question",
                   values_to = "response") %>%
      # Remove rows where both 'question' and 'response' are NA
      filter(!is.na(question) & !is.na(response)) %>%
      unite("question_scenario", scenario, question, sep = "_") %>%
      pivot_wider(names_from = question_scenario, values_from = response)
    
    # Append the cleaned data from this file to the main cleaned_data tibble
    # Ensure rows with NAs are retained by not filtering them out
    cleaned_data <- bind_rows(cleaned_data, temp_cleaned)
  } else {
    # If attention check fails, log the file name (optional)
    message(paste("Attention check failed for file:", file))
  }
}

# View the final combined cleaned_data tibble
print(cleaned_data)

# Define the output path for the final cleaned CSV file
output_path <- "../data/cleaned_data_final.csv"

# Save the final combined cleaned_data tibble as a CSV file
write_csv(cleaned_data, output_path)

```

#### Combined Data Processing

```{r}
# Load the combined dataset
dat <- read.csv("../data/quick_decisions_combined.csv")  # Read the dataset into a data frame

### Implement Exclusion Criteria: For participant with incomplete answers and lack of variance

# Process the dataset
cleaned_data <- dat %>%
  # Step 1: Filter out missing values
  mutate(is_complete = complete.cases(.)) %>%
  filter(complete.cases(.)) %>%
  
  # Step 2: Calculate variance for Justin and Nate
  rowwise() %>%
  mutate(
    justin_variance = var(c(2:11), na.rm = TRUE), # Variance for Justin
    nate_variance = var(c(12:22), na.rm = TRUE)          # Variance for Nate
  ) %>%
  ungroup() %>%
  
  # Step 3: Filter out participants based on variance criteria
  filter(!(justin_variance == 0 | nate_variance == 0)) %>%
  
  # Step 4: Drop intermediate columns used for exclusion
  select(-is_complete, -justin_variance, -nate_variance)

# Step 5: Calculate exclusion counts
  na_exclusions_count <- nrow(dat) - nrow(dat %>% filter(complete.cases(.))) # Count rows with missing values
  variance_exclusions_count <- nrow(dat %>% filter(complete.cases(.))) - nrow(cleaned_data) # Count participants excluded for lack of variance
  total_exclusions <- na_exclusions_count + variance_exclusions_count  # Calculate total exclusions

# Step 6: Create a summary table of exclusions
exclusions_summary <- tibble(
  Exclusion_Criteria = c(
    "Missing values (NA)",
    "Lack of variance",
    "Total exclusions"
  ),
  Count = c(
    na_exclusions_count,
    variance_exclusions_count,
    total_exclusions
  )
)

# Display the table
print(exclusions_summary)
```

#### Manipulation Check for Combined Dataset

Groups the data by condition and provides a clear comparison of how the participants perceived Justin's decision speed and Nate's decision speed across the two conditions. (Note: the mean should be different between Justin and Nate but it should be similar across condition).

Then we will determine whether there is a statistically significant difference in the mean scores of Justin's speed and Nate's speed.

```{r}
# Add Participant_ID column and make it the first column
cleaned_data <- cleaned_data %>%
  mutate(Participant_ID = row_number()) %>%
  select(Participant_ID, everything())

manipulation_stats <- cleaned_data %>%
  group_by(condition) %>%  # Group data by the 'condition' column
  summarize(
    mean_speed_justin = mean(Justin_Q1),  # Mean of Justin's speed
    mean_speed_nate = mean(Nate_Q1)     # Mean of Nate's speed
  )

# Paired t-test for Justin's and Nate's speeds for checking whether the manipulation was successful or not 
manipulation_combined_t_test <- t.test(
  cleaned_data$Justin_Q1, 
  cleaned_data$Nate_Q1, 
  paired = TRUE
)


print(manipulation_combined_t_test)
```

### Confirmatory analysis

Since the objective of this replication study was to closely mirror the original methodology, the statistical analysis replicated the 2x2 ANOVA test used in the original study. This statistical approach allowed for a direct assessment of the main effects of decision speed and moral condition on moral character evaluations, as well as their interaction, within a consistent analytical framework.

The 2x2 ANOVA was particularly appropriate here because it provides a straightforward method for comparing group means across these two factors, making it ideal for testing the original hypotheses about how decision speed and moral outcome influence moral judgments. By using the same statistical test as the original study, we ensured that any differences in results could be attributed to sample or context rather than methodological inconsistencies.

To enhance the robustness of our findings, we performed the analysis on both the combined dataset (aggregated across all three teams) and our team-specific dataset. The combined dataset allowed us to evaluate the generalizability of the findings across a more diverse sample while maintaining alignment with the target sample size of 300. The team-specific dataset, on the other hand, allowed us to ensure that any patterns observed in the combined analysis were consistent with those from our independently collected data.

While other methods, such as a linear mixed-effects model, could offer flexibility, the 2x2 ANOVA preserves the simplicity and interpretability central to the original study’s analysis. By using this approach, we stay true to the original design and ensure consistency in our replication study.

#### Confirmatory Analysis for Combined Dataset

```{r}
### Step 1: Calculate Average Scores for Justin and Nate

character_eval_combined <- cleaned_data %>% 
  mutate(
    # Calculate the average score for Justin across selected questions
    quick_justin_score = rowMeans(select(., Justin_Q2, Justin_Q3, Justin_Q4)),
    # Calculate the average score for Nate across selected questions
    slow_nate_score = rowMeans(select(., Nate_Q2, Nate_Q3, Nate_Q4))
  )

### Step 2: Summarize Average Scores by Condition

# Group data by condition for comparison of Justin's moral character evaluation and Nate's character evaluation across condition.

 character_eval_combined_stats <- character_eval_combined %>%
  group_by(condition) %>%  # Group data by the 'condition' column
  summarize(
    mean_justin = mean(quick_justin_score), # Mean of Justin's character evaluation
    mean_nate = mean(slow_nate_score)     # Mean of Nate's character evaluation
  )


### Step 3: Reshape Data for ANOVA
character_eval_combined_long <- character_eval_combined %>%
  pivot_longer(
    cols = c(quick_justin_score, slow_nate_score), # Select the new columns
    names_to = "speed",                           # Create a column for speed
    values_to = "score"                            # Create a column for scores
  ) %>%
  mutate(
    speed = ifelse(speed == "quick_justin_score", "Justin", "Nate") # Relabel speed column
  )  %>%
  select(Participant_ID, condition, speed, score) # Keep only essential columns



### Step 4: Run Repeated-Measures ANOVA
anova_combined <- aov(score ~ condition * speed + Error(Participant_ID/speed), data = character_eval_combined_long)
summary(anova_combined)
```

#### Confirmatory Analysis of Group 3's data

This analysis focuses on a subset of the data, specifically Group 3. We will perform the same 2 (moral condition: moral vs. immoral) x 2 (decision speed: quick vs. slow) ANOVA as in the full dataset but restricted to Group 3. This allows us to assess whether the observed effects are consistent within this specific subset.

```{r}
### Sub setting the data
g3_cleaned_data <- cleaned_data %>%
  filter(group == 3)


### Step 1: Calculate Average Scores for Justin and Nate
character_eval_group3 <- g3_cleaned_data %>% 
  mutate(
    # Calculate the average score for Justin across selected questions
    quick_justin_score = rowMeans(select(., Justin_Q2, Justin_Q3, Justin_Q4)),
    # Calculate the average score for Nate across selected questions
    slow_nate_score = rowMeans(select(., Nate_Q2, Nate_Q3, Nate_Q4))
  )

### Step 2: Summarize Average Scores by Condition

# Group data by condition for comparison of Justin's moral character evaluation and Nate's character evaluation across condition

 character_eval_group3_stats <- character_eval_group3 %>%
  group_by(condition) %>%  # Group data by the 'condition' column
  summarize(
    mean_justin = mean(quick_justin_score), # Mean of Justin's character evaluation
    mean_nate = mean(slow_nate_score)     # Mean of Nate's character evaluation
  )

### Step 3: Reshape Data for ANOVA
character_eval_group3_long <- character_eval_group3 %>%
  pivot_longer(
    cols = c(quick_justin_score, slow_nate_score), # Select the new columns
    names_to = "speed",                           # Create a column for speed
    values_to = "score"                            # Create a column for scores
  ) %>%
  mutate(
    speed = ifelse(speed == "quick_justin_score", "Justin", "Nate") # Relabel speed column
  )  %>%
  select(Participant_ID, condition, speed, score) # Keep only essential columns


### Step 4: Run Repeated-Measures ANOVA

anova_group3 <- aov(score ~ condition * speed + Error(Participant_ID/speed), data = character_eval_group3_long)
summary(anova_group3)
```

#### Visualization

```{r}
ggplot(character_eval_combined_long, aes(x = speed, y = score, fill = condition)) +
  geom_bar(
    stat = "summary",
    fun = "mean",
    position = position_dodge(width = 0.6),  # Columns closer together
    width = 0.5                             # Narrower columns
  ) +
  labs(
    title = "Moral Character Evaluations (Combined Data)",
    x = "Decision Speed (Justin / Nate)",
    y = "Mean Score (Likert Scale)",  # Updated y-axis label
    fill = "Moral Condition"          # Legend title for conditions
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 12),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),  # Centered title
    axis.text.x = element_text(size = 10),   # Adjust x-axis labels
    axis.title.y = element_text(size = 12, margin = margin(r = 15)),  # Add spacing to y-axis title
    axis.title.x = element_text(size = 12, margin = margin(t = 10))   # Add spacing to x-axis title
  ) +
  scale_fill_manual(
    values = c("moral" = "#4F81BD", "immoral" = "#9DC3E6"),  # Two shades for moral/immoral
    labels = c("moral" = "Moral", "immoral" = "Immoral")     # Custom legend labels
  ) +
  scale_y_continuous(
    limits = c(0, 7),  # Y-axis limits for Likert scale
    breaks = 1:7,      # Label y-axis by 1's
    expand = c(0, 0)   # No extra space above/below
  )
```

```{r}
ggplot(character_eval_group3_long, aes(x = speed, y = score, fill = condition)) +
  geom_bar(
    stat = "summary",
    fun = "mean",
    position = position_dodge(width = 0.6),  # Columns closer together
    width = 0.5                             # Narrower columns
  ) +
  labs(
    title = "Moral Character Evaluations (Group 3 Data)",
    x = "Decision Speed (Justin / Nate)",    # Updated x-axis label
    y = "Mean Score (Likert Scale)",         # Updated y-axis label
    fill = "Moral Condition"                # Legend title for conditions
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 12),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),  # Centered title
    axis.text.x = element_text(size = 10),   # Adjust x-axis labels
    axis.title.y = element_text(size = 12, margin = margin(r = 15)),  # Add spacing to y-axis title
    axis.title.x = element_text(size = 12, margin = margin(t = 10))   # Add spacing to x-axis title
  ) +
  scale_fill_manual(
    values = c("moral" = "#4F81BD", "immoral" = "#9DC3E6"),  # Two shades for moral/immoral
    labels = c("moral" = "Moral", "immoral" = "Immoral")     # Custom legend labels
  ) +
  scale_y_continuous(
    limits = c(0, 7),  # Y-axis limits for Likert scale
    breaks = 1:7,      # Label y-axis by 1's
    expand = c(0, 0)   # No extra space above/below
  )
```

# Exploratory Analysis

## Analysis on perceived certainty

The aim of this analysis is to perform a **2 (moral condition: moral vs. immoral) x 2 (decision speed: quick vs. slow)** repeated-measures ANOVA to compare group means for decision certainty evaluations of Justin and Nate.

Items accounting for decision certainty evaluation (Question 5-8):

5.  Would you say \[name\] was quite certain in his decision, or did \[name\] have hesitations about his decision? (1 = completely certain, 7 = considerable hesitations)
6.  How close do you think \[name\] was to choosing the alternate course of action? (1 = very close to, 7 = not close at all)
7.  How conflicted do you think \[name\] felt in making the decision? (1 = very conflicted, 7 = not at all conflicted)
8.  Based on the information provided, do you think \[name\] had many reservations about the decision? (1 = none at all, 7 = a whole lot)

```{r}
### Step 1: Calculate Average Scores for Justin and Nate

  # Since Questions 5 & 8 are reverse scored, we need to reverse code those items before we proceed with further analysis  Reverse coding is performed on a 1-7 scale to align with the scoring direction. 
  # High score = higher certainty on decision 
 
 #1a: Subset relevant columns

  certain_dat <- cleaned_data[, c(1:2, 7:10, 17:20, 23)]

  #1b: Reverse code Questions 5 and 8 for both Justin and Nate
  certain_dat$Justin_Q5R <- 8 - certain_dat$Justin_Q5  # Reverse code for Justin's Q5
  certain_dat$Justin_Q8R <- 8 - certain_dat$Justin_Q8  # Reverse code for Justin's Q8
  certain_dat$Nate_Q5R <- 8 - certain_dat$Nate_Q5      # Reverse code for Nate's Q5
  certain_dat$Nate_Q8R <- 8 - certain_dat$Nate_Q8      # Reverse code for Nate's Q8

  #1c: Calculate combined certainty scores 
  certain_eval_combined <- certain_dat %>%
    mutate(
      quick_justin_score = rowMeans(select(., Justin_Q5R, Justin_Q6, Justin_Q7, Justin_Q8R), na.rm = TRUE),
      slow_nate_score = rowMeans(select(., Nate_Q5R, Nate_Q6, Nate_Q7, Nate_Q8R), na.rm = TRUE)
    )

### Step 2: Summarize Average Scores by Condition
  
  # Group data by condition for comparison of Justin's decision certainty evaluation and Nate's decision certainty evaluation across condition

certain_eval_combined_stats <- certain_eval_combined %>%
  group_by(condition) %>%  # Group data by the 'condition' column
  summarize(
    mean_justin = mean(quick_justin_score), # Mean of Justin's character evaluation
    mean_nate = mean(slow_nate_score)     # Mean of Nate's character evaluation
  )

### Step 3: Reshape Data for ANOVA

certain_eval_combined_long <- certain_eval_combined %>%
  pivot_longer(
    cols = c(quick_justin_score, slow_nate_score), # Select the new columns
    names_to = "speed",                           # Create a column for speed
    values_to = "score"                            # Create a column for scores
  ) %>%
  mutate(
    speed = ifelse(speed == "quick_justin_score", "Justin", "Nate") # Relabel speed column
  )  %>%
  select(Participant_ID, condition, speed, score) # Keep only essential columns


### Step 4: Run Repeated-Measures ANOVA

anova_certain_combined <- aov(score ~ condition * speed + Error(Participant_ID/speed), data = certain_eval_combined_long)
summary(anova_certain_combined)
```

### Visualization

```{r}
ggplot(certain_eval_combined_long, aes(x = speed, y = score, fill = condition)) +
  geom_bar(
    stat = "summary",
    fun = "mean",
    position = position_dodge(width = 0.6),  # Columns closer together
    width = 0.5                             # Narrower columns
  ) +
  labs(
    title = "Decision Certainty Evaluations (Combined Data)",
    x = "Decision Speed (Justin / Nate)",    # Updated x-axis label
    y = "Mean Score (Likert Scale)",         # Updated y-axis label
    fill = "Moral Condition"                # Legend title for conditions
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 12),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),  # Centered title
    axis.text.x = element_text(size = 10),   # Adjust x-axis labels
    axis.title.y = element_text(size = 12, margin = margin(r = 15)),  # Add spacing to y-axis title
    axis.title.x = element_text(size = 12, margin = margin(t = 10))   # Add spacing to x-axis title
  ) +
  scale_fill_manual(
    values = c("moral" = "#4F81BD", "immoral" = "#9DC3E6"),  # Two shades for moral/immoral
    labels = c("moral" = "Moral", "immoral" = "Immoral")     # Custom legend labels
  ) +
  scale_y_continuous(
    limits = c(0, 7),  # Y-axis limits for Likert scale
    breaks = 1:7,      # Label y-axis by 1's
    expand = c(0, 0)   # No extra space above/below
  )

```

## Post-Hoc Comparison of all three 2x2 Anova Tests

### Post-Hoc Pairwise Comparisons on both Moral Character Evaluation & Decision Certainty Evaluation tests

#### Moral Character Evaluation

```{r}
# Combined Data: post hoc pairwise comparisons using estimated marginal means
posthoc_results <- emmeans(anova_combined, pairwise ~ condition * speed)
print(posthoc_results)

```

```{r}
# Group 3 Data: post hoc pairwise comparisons using estimated marginal means
posthoc_group3_results <- emmeans(anova_group3, pairwise ~ condition * speed)
print(posthoc_group3_results)
```

#### Decision Certainty Evaluation

```{r}
# Conduct post hoc pairwise comparisons using estimated marginal means
posthoc_certain_results <- emmeans(anova_certain_combined, pairwise ~ condition * speed)
print(posthoc_certain_results)
```

## Effect Size Calculation on both Moral Character Evaluation & Decision Certainty Evaluation tests

#### Moral Character Evaluation

```{r}
# Combined Data
# Refit the repeated measures model
anova_combined_lmer <- lmer(
  score ~ condition * speed + (1 | Participant_ID),
  data = character_eval_combined_long
)

# Calculate partial eta-squared for the mixed-effects model
eta2_combined <- eta_squared(anova_combined_lmer, partial = TRUE)

# Print results
print(eta2_combined)
```

```{r}
#Group 3 Data
# Refit the repeated measures model
anova_group3_lmer <- lmer(
  score ~ condition * speed + (1 | Participant_ID),
  data = character_eval_group3_long
)

# Calculate partial eta-squared for the mixed-effects model
eta2_group3 <- eta_squared(anova_group3_lmer, partial = TRUE)

# Print results
print(eta2_group3)
```

#### Decision Certainty Evaluation

```{r}
# Refit the repeated measures model
anova_certain_combined_lmer <- lmer(
  score ~ condition * speed + (1 | Participant_ID),
  data = certain_eval_combined_long
)

# Calculate partial eta-squared for the mixed-effects model
eta2_certain_combined <- eta_squared(anova_certain_combined_lmer, partial = TRUE)

# Print results
print(eta2_certain_combined)
```

### Interpretation

The analysis tested the hypothesis that individuals who make quick decisions about moral dilemmas are perceived as having stronger moral convictions than those who deliberate longer. A 2 (moral condition: moral vs. immoral) x 2 (decision speed: quick vs. slow) ANOVA was performed on group means of moral character evaluations in both the combined dataset and our group (3) dataset. For the combined dataset, there was a significant main effect of moral condition, with participants ranking moral decisions higher than immoral ones (F(1, 194) = 108.06, p < 2e-16). The pattern of interaction of moral condition and decision speed showed a significant interaction effect: moral character evaluations were generally amplified by decision speed, with quick decisions judged more favorably in moral scenarios and more harshly in immoral scenarios (F(1, 194) = 70.71, p = 8.83e-15). These findings align with the hypothesis that decision speed influences the strength of perceived moral convictions.

In the Group 3 dataset, similar patterns emerged, with a significant interaction effect between moral condition and decision speed (F(1, 184) = 92.63, p < .001). In moral contexts, quick decisions produced the highest evaluations of character, while slow judgments moderated evaluations within both moral and immoral contexts.

Additional exploratory analyses assessed perceived certainty as a covariate. These analyses revealed that faster decisions were related to greater perceived certainty (F(1, 408) = 218.43, p < 2e-16), further supporting the hypothesis that the speed of decision correlates with moral conviction. However, no significant main effect of moral condition (F(1, 408) = 0.275, p = 0.600) or interaction effect between condition and speed (F(1, 408) = 0.417, p = 0.519) was found. These results suggest that decision speed, rather than moral condition, drives perceptions of certainty.

## Discussion

Overall, this replication attempt validates the original study’s conclusion that quick decisions amplify perceptions of moral clarity and character, while slower decisions moderate these evaluations. This suggests that people interpret decision speed as an indicator of conviction, influencing how moral character is judged in various contexts.

Despite the successful replication, limitations include the use of an online sample via Prolific, which introduces variability compared to the original, in-person, university-based sample, and the simplified wallet scenario, which may not fully reflect the complexity of real-world moral dilemmas. Additionally, the experiment’s questions were verbose and highly similar, leading us to suspect —supported by variance tests— that participants were not fully reading and carefully weighing each question, which may have reduced the validity of the responses.

Future research could build on these findings by examining diverse moral dilemmas (as done in Experiment 2 of the original study), cultural differences in moral judgments, and the role of contextual factors such as stress and external pressures.

### Summary of Replication Attempt

Fortunately, the replication study conducted successfully reproduced the key findings of the original study: confirming that decision speed amplifies moral evaluations. As mentioned above, quick moral decisions were judged more favorably, while quick immoral decisions were judged more harshly, with slower decisions leading to more moderate evaluations. The experiment and analyses conducted adhered closely to the original methodology and extend its generalizability by recruiting a more diverse online participant sample. These results support the robustness of the original findings and highlight areas for further exploration: mainly, investigating more complex scenarios and polishing question design to enhance data quality. Overall, this replication underscores the reliability of the original study’s findings and highlights the importance of replication for advancing psychological research.

### Commentary

## Statement of Contributions

**Asad Tariq**: Conceptualization, Formal Analysis, Software, Methodology, Investigation, Data Curation, Writing

**Emily Han**: Conceptualization, Formal Analysis, Software, Methodology, Investigation, Data Curation, Writing

**Erika Garza-Elorduy**: Conceptualization, Formal Analysis, Investigation, Project Administration, Data Visualization, Writing

**Luna Bellitto**: Conceptualization, Formal Analysis, Investigation, Methodology, Project Administration, Writing

**Janna Wennberg**: Data Curation